### 2013-06-11 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
mc background edits and spell check
adding some example figures for domain decomposed monte carlo

\chapter{Parallel Monte Carlo Synthetic Acceleration}
\label{ch:parallel_methods}
For Monte Carlo methods, and in particular MCSA, to be viable at the
production scale, scalable parallel implementations are required. In a
general linear algebra context for discrete systems, this has not yet
been achieved. Therefore, we will develop and implement parallel
algorithms for these methods leveraging both the knowledge gained from
the general parallel implementations of Krylov methods and modern
parallel strategies for Monte Carlo as developed by the reactor
physics community. In order to formulate a parallel MCSA algorithm, we
recognize that the algorithm occurs in two stages, an outer iteration
performing fixed point iteration and applying the correction, and an
inner Monte Carlo solver that is generating the correction via the
adjoint or forward methods. The parallel aspects of both these
components must be considered.

In this chapter we briefly review particle transport methods for
domain decomposed Monte Carlo and then provide an analytic framework
from which to estimate its performance. We then devise an algorithm
based on these methods for efficient parallel solutions using the
Neumann-Ulam method and MCSA and verify its correctness by comparing
the numerical results against production Krylov methods. We then
perform parallel scaling studings to test its performance on both a
small computing cluster and a leadership class machine and compare
this performance to the production Krylov methods.

%%---------------------------------------------------------------------------%%
\section{Domain Decomposition for Monte Carlo}
\label{sec:msod}
As observed in the discussion on parallel Krylov methods in
Appendix~\ref{ch:linear_problem}, large-scale problems will surely
have their data partitioned such that each parallel process owns a
subset of the equations in the linear system. Given this convention,
the adjoint Monte Carlo algorithm must perform random walks over a
domain that is decomposed and must remain decomposed due to memory
limitations. This naturally leads us to seek parallel Monte Carlo
algorithms that handle domain decomposition.

In the context of radiation transport, Brunner and colleagues provided
a survey of algorithms for achieving this as implemented in production
implicit Monte Carlo codes \citep{brunner_comparison_2006}. In their
work they identify two data sets that are required to be communicated:
the sharing of particles that are transported from one domain to
another and therefore from one processor to another and a global
communication that signals if particle transport has been completed on
all processors. The algorithms presented are a fully-locking
synchronous scheme, an asynchronous-send/synchronous-receive pattern,
a traditional master/slave scheme, and a modified master/slave scheme
the implements a binary tree pattern for the global reduction type
operations needed to communicate between the master and slave
processes. They observed that the modified master/slave scheme
performed best in that global communications were implemented more
efficiently than those required by the asynchronous
scheme. Furthermore, none of these schemes handled load-imbalanced
cases efficiently. Such cases will be common if the source sampled in
the Monte Carlo random walk is not isotropic and not evenly
distributed throughout the global domain. It was noted that
efficiencies were improved by increasing the frequency by which
particle data was communicated between domain-adjacent
processors. However, this ultimately increases communication costs. In
2009, Brunner extended his work by using a more load-balanced approach
with a fully asynchronous communication pattern
\citep{brunner_efficient_2009}. Although the extended implementation
was more robust and allowed for scaling to larger numbers of
processors, performance issues were still noted with parallel
efficiency improvements needed in both the weak and strong scaling
cases for unbalanced problems. These results led Brunner to conclude
that a combination of domain decomposition and domain replication
could be used to solve some of these issues.

\subsection{Multiple-Set Overlapping-Domain Decomposition}
\label{subsec:msod}
In 2010, Wagner and colleagues developed the \textit{multiple-set
  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
applications for full-core light water reactor analysis
\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
their scheme employed similar parallel algorithms for particle
transport but a certain amount of overlap between adjacent domains was
used to decrease the number of particles leaving the local domain. In
addition, Wagner utilized a level of replication of the domain such
that the domain was only decomposed on $O(100)$ processors and if
replicated $O(1,000)$ times achieves simulation on $O(100,000)$
processors, thus providing spatial and particle parallelism. Each
collection of processors that constitutes a representation of the
entire domain is referred to as a set, and within a set overlap occurs
among its sub-domains. The original motivation was to decompose the
domain in a way that it remained in a physical cabinet in a large
distributed machine, thus reducing latency costs during
communication. A multiple set scheme is also motivated by the fact
that communication during particle transport only occurs within a set,
limiting communications during the transport procedure to a group of
$O(100)$ processors, a number that was shown to have excellent
parallel efficiencies in Brunner's work and therefore will scale well
in this algorithm. The overlapping domains within each set also
demonstrated reduced communication costs. On each processor, the
source is sampled in the local domain that would exist if no overlap
was used while tallies can be made over the entire overlapping domain.

%%---------------------------------------------------------------------------%%
\section{An Analytic Performance Framework for Domain-Decomposed Monte Carlo}
\label{sec:analytic_framework}
With Wagner's work \citep{wagner_hybrid_2010}, we observed that domain
overlap implemented with a domain decomposed Monte Carlo setting
reduces the amount of nearest neighbor communication that must occur
during transport. To date, parallel Neumann-Ulam methods have been
limited to full domain replication with parallelism exploited through
individual histories \citep{alexandrov_efficient_1998} and in this
chapter we will exploit Wagner's work to alleviate this. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progresses and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain. We wish
to characterize this communication not only because communication is
in general expensive, but also because these nearest-neighbor
communication sequences, specifically, have poor algorithmic strong
scaling \citep{gropp_high-performance_2001} in much the same way as a
matrix-vector multiply operation. Therefore, we desire a framework to
provide a simple, analytic theory based on the properties of the
linear system that will allow for estimates of the domain decomposed
behavior of the adjoint Neumann-Ulam method in terms of communication
costs.

\subsection{Leakage Fractions for Symmetric Systems}
\label{sec:symmetric_leakage_fractions}
When solving problems where the linear operator is symmetric, a host
of analytic theories exist based on the eigenvalue spectrum of the
operator that characterize their behavior in the context of
deterministic linear solvers. Using past work, these theories are
adapted to the domain decomposed adjoint Neumann-Ulam method using the
one-speed, two-dimensional neutron diffusion equation and spatial
discretization presented in \S~\ref{subsec:model_problem}. Using the
linear system generated by the discretization of the model problem, we
use a spectral analysis to generate analytic relations for the
eigenvalues of the operator based on system parameters. Using the
eigenvalue spectra, we then build relationships to characterize the
transport of stochastic histories in a decomposed domain and the
fraction of histories that leak from a domain and will therefore have
to be communicated. Finally, we compare these analytic results to
numerical experiments conducted with the model problem.

\subsubsection{Spectral Analysis}
\label{subsec:spectral_analysis}
The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the eigenvalues of the iteration matrix. We
will compute these eigenvalues by assuming eigenfunctions of the form
\citep{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these eigenfunctions should
give the eigenvalues of the matrix as they exist on the unit circle in
the complex plane.

\subsubsection{Iteration Matrix Spectrum}
\label{subsec:iteration_spectrum}
For the model problem, we first compute the eigenvalues for the
diffusion operator $\ve{D}$ by applying the operator to the
eigenfunctions and noting that $x=ih$ and $y=jh$:
\begin{multline}
  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
  \:.
  \label{eq:deriv_diff_1}
\end{multline}
Using Euler's formula, we can collapse the exponentials to
trigonometric functions:
\begin{equation}
  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
  \label{eq:deriv_diff_2}
\end{equation}

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, point Jacobi
preconditioning as outlined in \S\ref{sec:stochastic_preconditioning}
is sufficient to reduce the spectral radius of the iteration matrix
below unity and therefore ensure convergence of the Neumann
series. Applying this preconditioner, we are then solving the
following diffusion system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
operator with each row scaled by the diagonal component. As we have
defined a homogeneous domain, the scaling factor, $\alpha$, is the
same for all rows in the operator and defined as the $\phi_{i,j}$
coefficient from Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we then have the following spectrum of
preconditioned eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

@@

\subsubsection{Neumann Series Convergence}
\label{subsec:neumann_convergence}
As outlined in \S\ref{sec:mc_preliminaries}, the adjoint Monte Carlo
method is effectively an approximation to a stationary method. In the
adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
applications of the iteration matrix, are approximated by a random
walk of average length $k$ to yield the summation in
Eq~(\ref{eq:adjoint_neumann_solution})
\citep{dimov_new_1998,danilov_asymptotic_2000}. This random walk
length, or the number of transitions before the termination of a
history (either by the weight cutoff, absorption, or exiting the
global domain) is therefore approximately the number of stationary
iterations required to converge to the specified tolerance. In the
case of the adjoint Neumann-Ulam method, no such tolerance exists,
however, we have specified a weight cutoff, $W_c$, that determines
when low-weight histories will be prematurely terminated as their
contributions are deemed minute. After $k$ iterations, a stationary
method is terminated as the error has reached some fraction,
$\epsilon$, of the initial error:
\begin{equation}
  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm4}
\end{equation}
Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
Neumann-Ulam method, if we take this fraction to be the weight cutoff,
a measure of how accurately the contributions of a particular history
to the solution are tallied, we then have the following relationship
for $k$:
\begin{equation}
  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
  \label{eq:analytic_k}
\end{equation}
This then gives us a means to estimate the length of the random walks
that will be generated from a particular linear operator based on the
eigenvalues of its iteration matrix (independent of the linear
operator splitting chosen) and based on the weight cutoff parameter
used in the Neumann-Ulam method.

\subsubsection{Domain Leakage Approximations}
\label{subsubsec:domain_leak_approx}
In a domain decomposed situation, not all histories will remain within
the domain they started in and must instead be communicated. This
communication, expected to be expensive, was analyzed by Siegel and
colleagues for idealized, load balanced situations for full nuclear
reactor core Monte Carlo neutral particle simulations
\citep{siegel_analysis_2012}.  To quantify the number of particles
that leak out of the local domain they define a leakage fraction,
$\Lambda$, as:
\begin{equation}
  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
          \label{eq:leakage_fraction}
\end{equation}
For their studies, Siegel and colleagues assumed that the value of
$\Lambda$ was dependent on the total cross subsection of the system via
the Wigner rational approximation. Outlined more thoroughly by Hwang's
chapter in \citep{azmy_nuclear_2010}, we will use both the Wigner
rational approximation and the mean chord approximation as a means to
estimate the leakage fraction.

@@

\subsubsection{Numerical Experiments}
\label{subsec:numerical_experiments}
To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest eigenvalue of the diffusion operator
using an iterative eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
absorption cross varied from 0 to 100 while the scattering cross
subsection was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
the measured spectral radius of the iteration matrix and the computed
spectral radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative eigenvalue
solver.
\begin{figure}[t!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=5in,clip]{chapters/parallel_mc/measured_spec_rad.png}
    \end{center}
    \caption{\textbf{Measured and analytic preconditioned diffusion
        operator spectral radius as a function of the absorption cross
        subsection to scattering cross subsection ratio.} \textit{Values of
        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
        computed numerically by an eigensolver while the black dashed
        data was generated by Eq~(\ref{eq:iteration_radius}).}}
    \label{fig:measured_spec_rad}
  \end{spacing}
\end{figure}

@@

\subsection{Communication Costs for Symmetric Systems}
\label{sec:symmetric_comm_costs}
Use Seigel's work to get communication costs.

%%---------------------------------------------------------------------------%%
\section{Fully Asynchronous MSOD Neumann-Ulam Algorithm}
\label{sec:asynchronous_algorithm}
We can take much of what was learned from Brunner and Wagner's
parallel Monte Carlo methods for radiation transport and directly
apply it to a parallel formulation of our stochastic linear
solvers. Direct analogs can be derived from these works by noting that
the primary difference between solving a linear system with Monte
Carlo methods and fixed source Monte Carlo transport problems is the
content of the Markov chains that are generated. The transitions
represented by these chains are bound by probabilities and weights and
are initiated by the sampling of a source. In the context of transport
problems, those transitions represent events such as particle
scattering and absorption with probabilities that are determined by
physical data in the form of cross sections. For stochastic matrix
inversion, those transitions represent moving between the equations of
the linear system (and therefore the physical domain which they
represent) and their probabilities are defined by the coefficients of
those equations. Ultimately, we tally the contributions to generate
expectation values in the desired states as we progress through the
chains. Therefore, parallel methods for Monte Carlo radiation
transport can be abstracted and we can use those concepts that apply
to matrix inversion methods as an initial means of developing a
parallel Neumann-Ulam-type solver. In this section the fully
asynchronous MSOD Neumann-Ulam algorithm developed for this work is
presented step-by-step in detail.

\subsection{Parallel Random Number Generation}
\label{subsec:parallel_rng}
Correct parallel Monte Carlo requires parallel random number
generation. To ensure distinct Markov chains for each history in the
problem, regardless of its state in phase space or parallel space, a
unique stream of random numbers is necessary to achieve uncorrelated
tally results for the solution.

\subsection{Generating the Transport Domain with Overlap and Preconditioning}
\label{subsec:domain_generation}

@@

\subsection{Transition Processing Kernel}
\label{subsec:transition_processing}
Processing the transition of a single stochastic history from one
state to another in the system is the fundamental kernel of the
transport procedure. In this kernel, an incoming history has a state
consisting of the current discrete state in the system in which it
resides and its current weight and is processed as given by
Algorithm~\ref{alg:transition_process}.
\begin{algorithm}[h!]
  \caption{Stochastic History Transition Processing Kernel}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State processTransition( history ):
    \State i = history.state
    \State j = sampleRowCDF( C[i] )
    \State history.state = j
    \State history.weight *= w[i,j]
  \end{algorithmic}
\end{algorithm}
The core component here is the sampling of the CDF for a particular
state generated by the probablities from the Neumann-Ulam
decomposition. Upon exiting, the kernel has modified the state of the
history with a new state and multiplied the weight with the
corresponding transition weight to form the product in
Eq~(\ref{eq:direct_permutation_weight}).

@@

\subsection{Multiple Set Algorithm}
\label{subsec:multiple_sets_algorithm}

@@

\subsection{Multiple Set Domain Decomposition}
\label{subsec:multiple_set_decomposition}

\subsubsection{Weak Scaling}
\label{subsubsec:multiple_set_weak}

>>>>

\chapter{Parallel Monte Carlo Synthetic Acceleration}
\label{ch:parallel_methods}
For MCSA to be viable at the production scale, scalable parallel
implementations of the algorithm are required. Reviewing the
literature, MCSA has yet to be parallelized and the Neumann-Ulam Monte
Carlo method on which it is built has only been parallelized through
history-level parallelism with full domain replication. In order to
solve large linear systems with MCSA, a domain-decomposed parallel
strategy is required. In the formulation of a parallel MCSA algorithm,
we recognize that the algorithm occurs in two stages, an outer
iteration performing fixed point iteration and applying the
correction, and an inner Monte Carlo solver that is generating the
correction via the adjoint or forward methods. The parallel aspects of
both these components must be considered. Therefore, we will develop
and implement parallel algorithms for the Neumann-Ulam and MCSA
methods leveraging both the knowledge gained from the general parallel
implementations of Krylov methods reviewed in
Appendix~\ref{ch:linear_problem} and modern parallel strategies for
domain decomposed Monte Carlo as developed by the reactor physics
community.

In this chapter we briefly review particle transport methods for
domain decomposed Monte Carlo. Considering a domain decomposed
Neumann-Ulam method, we derive an analytic framework based on
algebraic quantities from which to estimate its performance when
applied to linear systems. We then devise a parallel algorithm for the
Neumann-Ulam method based on the multiple-set overlapping-domain
decomposition algorithm and a parallel algorithm for the MCSA that
leverages the parallel Monte Carlo algorithm and general parallel
matrix-vector operations. In identical fashion to the serial
computations presented in Chapter~\ref{ch:spn_equations}, we use
parallel solutions for the $SP_N$ equations for the fuel assembly
criticality problem to verify the correctness of the parallel
algorithm and implementation by comparing the numerical results
against production Krylov methods. Using the verified implementation
of the algorithm, we perform parallel scaling studings to test its
performance on both a small computing cluster and a leadership class
machine and compare this performance to the production Krylov methods.

%%---------------------------------------------------------------------------%%
\section{Domain Decomposed Monte Carlo}
\label{sec:msod}
As observed in the discussion on parallel Krylov methods in
Appendix~\ref{ch:linear_problem}, large-scale problems will surely
have their data partitioned such that each parallel process owns a
subset of the equations in the linear system. Given this convention,
the Neumann-Ulam Monte Carlo algorithm must perform random walks over
a domain that is decomposed and must remain decomposed due to memory
limitations. This naturally leads us to seek parallel Monte Carlo
algorithms that handle domain decomposition.

\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/ddmc_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Domain decomposed Monte Carlo transport example
      illustrating how domain decomposition requires parallel
      communication of particles.}}
  \label{fig:ddmc_example}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/ddnu_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Domain decomposed Neumann-Ulam example illustrating
      how domain decomposition requires parallel communication of
      histories.} \textit{Each mesh point corresponds to an equation
      in the linear system and the coupling among equations is
      described by the discretization of the problem.}}
  \label{fig:ddnu_example}
\end{figure}

\clearpage

%%---------------------------------------------------------------------------%%
\section{Analytic Performance Framework for Domain-Decomposed Monte Carlo}
\label{sec:analytic_framework}
To date, parallel Neumann-Ulam methods have been limited to full
domain replication with parallelism exploited through individual
histories \citep{alexandrov_efficient_1998} and in this chapter we
will exploit particle transport algorithms to alleviate this. To
accomplish this, we recognize from the literature that stochastic
histories must be transported from domain to domain as the simulation
progresses and they transition to states that are not in the local
domain. Because we have chosen a domain decomposition strategy in a
parallel environment, this means that communication of these histories
must occur between compute nodes owning neighboring pieces of the
global domain. We wish to characterize this communication not only
because communication is in general expensive, but also because these
nearest-neighbor communication sequences, specifically, have poor
algorithmic strong scaling \citep{gropp_high-performance_2001} in much
the same way as a matrix-vector multiply operation. Therefore, we
desire a framework to provide a simple, analytic theory based on the
properties of the linear system that will allow for estimates of the
domain decomposed behavior of the Neumann-Ulam method in terms of the
amount of information that must be communicated.

When solving problems where the linear operator is symmetric, a host
of analytic theories exist based on the eigenvalue spectrum of the
operator that characterize their behavior in the context of
deterministic linear solvers. Using past work, these theories are
adapted to the domain decomposed Neumann-Ulam method using the
one-speed, two-dimensional neutron diffusion equation and spatial
discretization presented in
Appendix~\ref{chap:diffusion_problem}. Using the linear system
generated by the discretization of the model problem, we use a
spectral analysis to generate analytic relations for the eigenvalues
of the operator based on system parameters. Using the eigenvalue
spectra, we then build relationships to characterize the transport of
stochastic histories in a decomposed domain and the fraction of
histories that leak from a domain and will therefore have to be
communicated. Finally, we compare these analytic results to numerical
experiments conducted with the model problem.

\subsection{Spectral Analysis}
\label{subsec:spectral_analysis}
The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the eigenvalues of the iteration matrix. We
will compute these eigenvalues by assuming eigenfunctions of the form
\citep{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these eigenfunctions should
give the eigenvalues of the matrix as they exist on the unit circle in
the complex plane.

For the model problem, we first compute the eigenvalues for the
diffusion operator $\ve{D}$ by applying the operator to the
eigenfunctions and noting that $x=ih$ and $y=jh$:
\begin{multline}
  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
  \:.
  \label{eq:deriv_diff_1}
\end{multline}
Using Euler's formula, we can collapse the exponentials to
trigonometric functions:
\begin{equation}
  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
  \label{eq:deriv_diff_2}
\end{equation}

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, point Jacobi
preconditioning as outlined in
\S~\ref{subsec:stochastic_preconditioning} is sufficient to reduce the
spectral radius of the iteration matrix below unity and therefore
ensure convergence of the Neumann series. Applying this
preconditioner, we are then solving the following diffusion system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
operator with each row scaled by the diagonal component. As we have
defined a homogeneous domain, the scaling factor, $\alpha$, is the
same for all rows in the operator and defined as the $\phi_{i,j}$
coefficient from Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we then have the following spectrum of
preconditioned eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

@@

\subsection{Neumann Series Convergence}
\label{subsec:neumann_convergence}
As outlined in \S\ref{sec:mc_preliminaries}, the adjoint Monte Carlo
method is effectively an approximation to a stationary method. In the
adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
applications of the iteration matrix, are approximated by a random
walk of average length $k$ to yield the summation in
Eq~(\ref{eq:adjoint_neumann_solution})
\citep{dimov_new_1998,danilov_asymptotic_2000}. This random walk
length, or the number of transitions before the termination of a
history (either by the weight cutoff, absorption, or exiting the
global domain) is therefore approximately the number of stationary
iterations required to converge to the specified tolerance. In the
case of the adjoint Neumann-Ulam method, no such tolerance exists,
however, we have specified a weight cutoff, $W_c$, that determines
when low-weight histories will be prematurely terminated as their
contributions are deemed minute. After $k$ iterations, a stationary
method is terminated as the error has reached some fraction,
$\epsilon$, of the initial error:
\begin{equation}
  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm4}
\end{equation}
Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
Neumann-Ulam method, if we take this fraction to be the weight cutoff,
a measure of how accurately the contributions of a particular history
to the solution are tallied, we then have the following relationship
for $k$:
\begin{equation}
  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
  \label{eq:analytic_k}
\end{equation}
This then gives us a means to estimate the length of the random walks
that will be generated from a particular linear operator based on the
eigenvalues of its iteration matrix (independent of the linear
operator splitting chosen) and based on the weight cutoff parameter
used in the Neumann-Ulam method.

\subsection{Domain Leakage Approximations}
\label{subsec:domain_leak_approx}
In a domain decomposed situation, not all histories will remain within
the domain they started in and must instead be communicated. This
communication, expected to be expensive, was analyzed by Siegel and
colleagues for idealized, load balanced situations for full nuclear
reactor core Monte Carlo neutral particle simulations
\citep{siegel_analysis_2012}.  To quantify the number of particles
that leak out of the local domain they define a leakage fraction,
$\Lambda$, as:
\begin{equation}
  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
          \label{eq:leakage_fraction}
\end{equation}
For their studies, Siegel and colleagues assumed that the value of
$\Lambda$ was dependent on the total cross subsection of the system via
the Wigner rational approximation. Outlined more thoroughly by Hwang's
chapter in \citep{azmy_nuclear_2010}, we will use both the Wigner
rational approximation and the mean chord approximation as a means to
estimate the leakage fraction.

@@

\subsection{Numerical Experiments}
\label{subsec:numerical_experiments}
To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest eigenvalue of the diffusion operator
using an iterative eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
absorption cross varied from 0 to 100 while the scattering cross
subsection was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
the measured spectral radius of the iteration matrix and the computed
spectral radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative eigenvalue
solver.
\begin{figure}[t!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=5in,clip]{chapters/parallel_mc/measured_spec_rad.png}
    \end{center}
    \caption{\textbf{Measured and analytic preconditioned diffusion
        operator spectral radius as a function of the absorption cross
        subsection to scattering cross subsection ratio.} \textit{Values of
        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
        computed numerically by an eigensolver while the black dashed
        data was generated by Eq~(\ref{eq:iteration_radius}).}}
    \label{fig:measured_spec_rad}
  \end{spacing}
\end{figure}

@@

%%---------------------------------------------------------------------------%%
\section{Fully Asynchronous Neumann-Ulam Algorithm}
\label{sec:asynchronous_algorithm}

In the context of radiation transport, in 2009 Brunner and colleagues
provided a fully asynchronous domain decomposed parallel aglorithm as
implemented in production implicit Monte Carlo codes
\citep{brunner_efficient_2009}. In their work they identify two data
sets that are required to be communicated: the sharing of particles
that are transported from one domain to another and therefore from one
processor to another and a global communication that signals if
particle transport has been completed on all processors. Although the
implementation was robust and allowed for scaling to large numbers of
processors, performance issues were still noted with parallel
efficiency improvements needed in both the weak and strong scaling
cases for unbalanced problems. These results led Brunner to conclude
that a combination of domain decomposition and domain replication
could be used to solve some of these issues.

We can take much of what was learned from Brunner's parallel Monte
Carlo method for radiation transport and directly apply it to a
parallel formulation of the Neumann-Ulam method. Direct analogs can be
derived from these works by noting that the primary difference between
solving a linear system with Monte Carlo methods and fixed source
Monte Carlo transport problems is the content of the Markov chains
that are generated. The transitions represented by these chains are
bound by probabilities and weights and are initiated by the sampling
of a source. In the context of transport problems, those transitions
represent events such as particle scattering and absorption with
probabilities that are determined by physical data in the form of
cross sections. For stochastic matrix inversion, those transitions
represent moving between the equations of the linear system (and
therefore the physical domain which they represent) and their
probabilities are defined by the coefficients of those
equations. Ultimately, we tally the contributions to generate
expectation values in the desired states as we progress through the
chains. Therefore, parallel methods for Monte Carlo radiation
transport can be abstracted and we can use those concepts that apply
to matrix inversion methods as an initial means of developing a
parallel Neumann-Ulam-type solver. In this section the fully
asynchronous MSOD Neumann-Ulam algorithm developed for this work is
presented step-by-step in detail.

\subsection{Generating the Transport Domain with Preconditioning}
\label{subsec:domain_generation}

@@

\subsection{Transition Processing Kernel}
\label{subsec:transition_processing}
Processing the transition of a single stochastic history from one
state to another in the system is the fundamental kernel of the
transport procedure. In this kernel, an incoming history has a state
consisting of the current discrete state in the system in which it
resides and its current weight and is processed as given by
Algorithm~\ref{alg:transition_process}.
\begin{algorithm}[h!]
  \caption{Stochastic History Transition Processing Kernel}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State processTransition( history ):
    \State i = history.state
    \State j = sampleRowCDF( C[i] )
    \State history.state = j
    \State history.weight *= w[i,j]
  \end{algorithmic}
  \label{alg:transition_process}
\end{algorithm}
The core component here is the sampling of the CDF for a particular
state generated by the probablities from the Neumann-Ulam
decomposition. Upon exiting, the kernel has modified the state of the
history with a new state and multiplied the weight with the
corresponding transition weight to form the product in
Eq~(\ref{eq:direct_permutation_weight}).

@@

%%---------------------------------------------------------------------------%%
\subsection{Multiple-Set Overlapping-Domain Neumann-Ulam Algorithm}
\label{subsec:msod}
In 2010, Wagner and colleagues developed the \textit{multiple-set
  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
applications for full-core light water reactor analysis
\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
their scheme employed similar parallel algorithms for particle
transport but a certain amount of overlap between adjacent domains was
used to decrease the number of particles leaving the local domain. In
addition, Wagner utilized a level of replication of the domain such
that the domain was only decomposed on $O(100)$ processors and if
replicated $O(1,000)$ times achieves simulation on $O(100,000)$
processors, thus providing spatial and particle parallelism. Each
collection of processors that constitutes a representation of the
entire domain is referred to as a set, and within a set overlap occurs
among its sub-domains. The original motivation was to decompose the
domain in a way that it remained in a physical cabinet in a large
distributed machine, thus reducing latency costs during
communication. A multiple set scheme is also motivated by the fact
that communication during particle transport only occurs within a set,
limiting communications during the transport procedure to a group of
$O(100)$ processors, a number that was shown to have excellent
parallel efficiencies in Brunner's work and therefore will scale well
in this algorithm. The overlapping domains within each set also
demonstrated reduced communication costs. On each processor, the
source is sampled in the local domain that would exist if no overlap
was used while tallies can be made over the entire overlapping domain.

\subsection{Generating Overlap}
\label{subsec:msod_overlap}

\subsection{Generating Multiple Sets}
\label{subsec:msod_sets}

\subsection{MSOD Neumann Ulam Algorithm}
\label{subsec:msod_algorithm}

@@
